---
title: "Assignment 2 - MD Fardin"
author: "MD Fardin"
date: "`r Sys.Date()`"
output: html_document
---

### MD Fardin

### Data 710 - Basic Analytics

### Professor Cohen

### 11/08/2024

#### Starting the session:

```{r include=FALSE}
# Start Session
rm(list = ls())
gc()

# Setting the Working Directory
setwd("D:/Graduate School Files/Data 710 -  Basic Analytics/Assignment 2")

# Load Packages
library(readxl)
library(ggplot2)
library(dplyr)
library(sf)
library(sp)
library(tidyverse)
library(descr) 
library(leaflet)
library(ggthemes)
library(writexl) 
```

#### Data Wrangling:

#### Data Wrangling for Demography Table:

```{r include=FALSE}
# Load the Data
Demography <- read_xlsx("Demography Data 2021-23.xlsx") # Demography Data
Weight <- read_xlsx("Weight Health Questionaire 2021-23.xlsx") # Weight Data

names(Demography)
```

```{r include=FALSE}
# I will only change the names for the columns that I need. 
names(Demography)[c(4, 5, 8, 11, 12, 14, 15, 16)] <- paste(c("Gender", "Age", "Race", "US Army Duty", "Country of Birth", "Education Level", "Marital Status", "Pregnancy Status"))

head(Demography)

names(Demography)
```

```{r include=FALSE}
# Not all variables are needed from the data, so I want a new table with the variables I need for this analysis
Demography2 <- Demography[, c("SEQN", "Gender", "Age", "Race", "US Army Duty", "Country of Birth", "Education Level", "Pregnancy Status")]

Demography2
```

```{r include=FALSE}
# Recode Variables for Demography2
Demography2$Gender <- ifelse(Demography2$Gender == 1, "Male", "Female")

Demography2$Race <- factor(Demography2$Race,
                           levels = c(1, 2, 3, 4, 6, 7),
                           labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Asian", "Other Race - Including Multi-Racial"))

Demography2$`US Army Duty` <- factor(Demography2$`US Army Duty`,
                                     levels = c(1, 2, 7, "NA"),
                                     labels = c("Yes", "No", "Refused", "Missing"))

Demography2$`Country of Birth` <- factor(Demography2$`Country of Birth`,
                                         levels = c(1, 2, "NA"),
                                         labels = c("Born in 50 US states or Washington", "Others", "Missing"))

Demography2$`Education Level` <- factor(Demography2$`Education Level`,
                                        levels = c(1, 2, 3, 4, 5, 9, "NA"),
                                        labels = c("Less than 9th grade", "9-11th grade", "High school graduate/GED or equivalent", "Some college or AA degree", "College graduate or above", "Don't know", "Missing"))

Demography2$`Pregnancy Status` <- factor(Demography2$`Pregnancy Status`,
                                         levels = c(1, 2, 3, "NA"),
                                         labels = c("Yes", "Not Pregnant", "Doesn't Know", "Missing"))
```

#### Data Wrangling for Weight Table:

```{r include=FALSE}
names(Weight)
```

```{r include=FALSE}
# Recode 
names(Weight)[c(2, 3, 4, 5)] <- paste(c("Height (Inches)", "Weight (Pounds)", "Weight (Last year;Pounds)", "Tried Weight Loss"))

head(Weight)

names(Weight)
```

```{r include=FALSE}
# Copy of data for data analysis for variable recode
Weight2 <- Weight
```

```{r include=FALSE}
# Recode Variables in the Weight2 table where I keep the numeric values as numbers while turning, "9999" and "7777" and "NA" to their string value:

Weight2$`Height (Inches)`<- ifelse(
  Weight2$`Height (Inches)` == 7777, "Refused",
  ifelse(
    Weight2$`Height (Inches)` == 9999, "Don't Know",
    ifelse(is.na(Weight2$`Height (Inches)`), "Missing",
           as.character(Weight2$`Height (Inches)`)
    )
  )
)


Weight2$`Weight (Pounds)`<- ifelse(
  Weight2$`Weight (Pounds)` == 7777, "Refused",
  ifelse(
    Weight2$`Weight (Pounds)` == 9999, "Don't Know",
    ifelse(is.na(Weight2$`Weight (Pounds)`), "Missing",
           as.character(Weight2$`Weight (Pounds)`)
    )
  )
)

Weight2$`Weight (Last year;Pounds)`<- ifelse(
  Weight2$`Weight (Last year;Pounds)` == 7777, "Refused",
  ifelse(
    Weight2$`Weight (Last year;Pounds)` == 9999, "Don't Know",
    ifelse(is.na(Weight2$`Weight (Last year;Pounds)`), "Missing",
           as.character(Weight2$`Weight (Last year;Pounds)`)
    )
  )
)

Weight2$`Tried Weight Loss` <- factor(Weight2$`Tried Weight Loss`,
                                     levels = c(1, 2, 7, 9, "NA"),
                                     labels = c("Yes", "No", "Refused", "Don't Know", "Missing"))
```

### Creating a new Variable to calculate BMI

```{r include=FALSE}
# Creating a BMI Variable

Weight2 <- Weight2 %>%
  # Filters rows with non-numeric values for height and weight
  filter(!`Height (Inches)` %in% c("Refused", "Don't know", "Missing"),
         !`Weight (Pounds)` %in% c("Refused", "Don't know", "Missing")) %>%
  # Convert height and weight to numeric
  mutate(
    `Height (Inches)` = as.numeric(`Height (Inches)`),
    `Weight (Pounds)` = as.numeric(`Weight (Pounds)`),
    # Calculate BMI
    BMI = ((`Weight (Pounds)`) / (`Height (Inches)` ^ 2) * 703)
  )
```

#### BMI Categories Indicator [(According to the National Heart, Lung, and Blood Institute (nhlbi)](https://www.nhlbi.nih.gov/health/educational/lose_wt/BMI/bmicalc.htm):

```{r include=FALSE}
# Creating a BMI Indicator
Weight2 <- Weight2 %>%
  mutate(
    `BMI Indicator` = case_when(
      BMI < 18.5 ~ "Underweight",
      BMI >= 18.5 & BMI < 25 ~ "Normal weight",
      BMI >= 25 & BMI < 30 ~ "Overweight",
      BMI >= 30 ~ "Obesity",
      TRUE ~ NA_character_  # Handles any unexpected NA or non-numeric BMI values
    )
  )
```

### Merging the Data

```{r include=FALSE}
# I will merge both the wrangled tables into one:
DATA <- merge(Demography2, Weight2, by = "SEQN", all.x = TRUE)
head(DATA)
```

```{r include=FALSE}
# Making a copy of my DATA table in case of issues. 
DATA2 <- DATA
```

### More Wrangling:

```{r include=FALSE}
# Removing anyone under the age of 18
DATA2 <- DATA2 %>%
  filter(Age >= 18)

# Removing any "NA" from "BMI Indicator" variable:
DATA2 <- DATA2 %>%
  filter(!is.na(`BMI Indicator`))

# Removing any "Don't Know" and "Missing" from the "Education Level" Variable:
DATA2 <- DATA2 %>%
  filter(!`Education Level` %in% c("Don't Know", "Missing"))

# Removing anyone who is Pregnant:
DATA2 <- DATA2 %>%
  filter(`Pregnancy Status` != "Yes")

# Females who don't know if they're pregnant
f_unknown_pregnancy <- DATA2 %>%
  filter(Gender == "Female" & `Pregnancy Status` == "Doesn't Know") %>%
  nrow()

f_unknown_pregnancy

# Out of curiosity, how many men selected "Doesn't Know" for the Pregnancy question
m_idk_pregnancy <- DATA2 %>%
  filter(Gender == "Male" & `Pregnancy Status` == "Doesn't Know") %>%
  nrow()

m_idk_pregnancy
```

### Analyze the Data:

```{r include=FALSE}
# The number of people in this sample
nrow(DATA2)

# BMI indicator count
freq(DATA2$`BMI Indicator`)

# Education Level
freq(DATA2$`Education Level`)

# BMI statistics
summary(DATA2$BMI)

# Weight Statistics
summary(DATA2$`Weight (Pounds)`)

# BMI Indicator by Education Level
BMI_Education <- table(DATA2$`Education Level`, DATA2$`BMI Indicator`)

#BMI Indicator by Education Level proportion
proptable_bmiedu <- prop.table(BMI_Education)
```

```{r include=FALSE}
# A bar graph for the BMI Indication
ggplot(DATA2, aes(x = `Education Level`, fill = `BMI Indicator`)) +
  geom_bar(position = "dodge") +
  labs(
    title = "BMI Indicator by Education Level",
    x = "Education Level",
    y = "# of People",
    fill = "BMI Indicator"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r include=FALSE}
# Correlation Between Education Level and BMI
kruskal_bmi <- kruskal.test(BMI ~ `Education Level`, data = DATA2)
kruskal_bmi
```

```{r include=FALSE}
# How many people in the sample tried weight loss
freq(DATA2$`Tried Weight Loss`)

# How many people in each indicator tried weight loss
table(DATA2$`BMI Indicator`, DATA2$`Tried Weight Loss`)

# Proportion of the sample that tried weight loss
prop.table(table(DATA2$`BMI Indicator`, DATA2$`Tried Weight Loss`))
```

#### Saving my table as an Excel file:

```{r include=FALSE}
# Saving DATA2 in my working directory 
write_xlsx(DATA2, path = "DATA2.xlsx")
```
